---
title: 'Data Analysis Assignment #1 (50 points total)'
author: Russell, Jamia
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```
-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Whole (skewness=.047), Shuck
(skewness=0.64), Ratio (skewness=.71), Volume (skewness=.44), and Rings (skewness=1.24) are all positively skewed, which suggests the presence of high outliers causing the mean to be greater than the median. Height has a closely symmetrical normal distribution. Negatively skewed are Length and Diameter. 

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
t <- table(mydata$SEX, mydata$CLASS)
addmargins(t)

freq <- data.frame(t)
colnames(freq) <- c('Sex', 'Class', 'Frequency')
ggplot(freq, aes(x=Class, y=Frequency, fill=Sex, group=Sex)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  ggtitle("Abalone by Class & Sex") 
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: I would not have anticipated the number of infants in classes A4 and A5. As they are supposed to be older classes of Abalone. It is unclear if there is an issue with the classification (unclear, not able to tell) of the abalones or just improper designation.

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
sm<-sample(seq(1:1036),200)
work<-mydata[sm,]
plot(work[,2: 6],cex=.5,lwd=.3)
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Whole Weight & Volume')) +
  scale_color_brewer(palette="Set3") +
  ylab("Whole Weight - Grams") +
  xlab(bquote('Volume ('~cm^3~')'))
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1,) +
  scale_color_brewer(palette="Pastel2") +
  ggtitle('Shuck Weight & Whole Weight') +
  geom_abline(intercept=0, slope=max(mydata$SHUCK/mydata$WHOLE), size=.25, linetype=2) +
  ylab("Shuck Weight - Grams") +
  xlab("Whole Weight - Grams") 
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: Variability seems to be more present in plot 2(a) compared to 2(b). It seems that weight and volume, carry a weaker relationship than shuck weight and whole weight.As shown in A5 abalone weight does increase as age does. 

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a,fig.height=7,fig.width=7}
histo <- ggplot(mydata, aes(RATIO, fill=SEX)) +
  facet_grid(. ~ SEX, space="free") +
  geom_histogram(bins=14, color="black", size=.2) + 
  ggtitle("Ratio Histograms") +
  scale_color_brewer(palette="Set3") +
  theme(legend.position="none", axis.title.x=element_blank()) +
  ylab("Frequency")
  
boxx <- ggplot(mydata, aes(y=RATIO, fill=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_boxplot(color="black", size=.2, outlier.size=2,  outlier.shape=1) + 
    ggtitle("Ratio Boxplots") +
  scale_color_brewer(palette="Set3") +
  theme(legend.position="none", axis.title=element_blank())

qoq <- ggplot(mydata, aes(sample=RATIO, color=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_qq(cex=2, shape=1) + 
  ggtitle("Ratio QQ Plots") +
  scale_color_brewer(palette="Set3") +
  geom_qq_line(color="black", cex=.5) +
  theme(legend.position="none") 
  
grid.arrange(histo, boxx, qoq, nrow=3)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: The displays show a non-normal distribution that are skewing to the right. Female and infant skews are more positive than male skews, which may be causing the shift in distribution.

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
infant_o <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
female_o <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
male_o <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(infant_o, female_o, male_o))
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: Extreme outliers come from the female and infant class. Infants outliers are primarily from the shuck/volume ratio and they are coming from earlier classes A1-A3. 

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
boxx_v <- ggplot(mydata, aes(x=CLASS, y=VOLUME, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="Pastel2") +
  ggtitle("Volume & Class") +
  theme(legend.position="none")

boxx_w <- ggplot(mydata, aes(x=CLASS, y=WHOLE, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="Pastel2") +
  ggtitle("Whole Weight & Class") +
  theme(legend.position="none")

scatt_v <- ggplot(data = mydata, aes(x = RINGS, y = VOLUME)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="Pastel2") +
  ggtitle("Volume & Rings") +
  theme(legend.position="none")

scatt_w <- ggplot(data = mydata, aes(x = RINGS, y = WHOLE)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="Pastel2") +
  ggtitle("Whole Weight & Rings") +
  theme(legend.position="none")

grid.arrange(boxx_v, boxx_w, scatt_v, scatt_w, nrow=2)
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer:The variable combinations are not definitive. They do not adequately predict the age of the abalones. Volume and weight do have a positive correlation with the number of rings increasing but no relation to age. Similarly, 
volume and class and wight and class display higher volume or weight are most often in a higher class of age. However, this correlation exists across three different classes - A3, A4, and A5. Weight is either a poor indicator of age or that there is an error in designation/classification.

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
agg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))

mean_vol <- matrix(agg$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

mean_shuck <- matrix(agg$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

mean_ratio <- matrix(agg$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

kable(mean_vol, caption="Volume")
kable(mean_shuck, caption="Shuck")
kable(mean_ratio, caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
colnames(agg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')
cr <- ggplot(data = agg, aes(x = Class, y = Ratio, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Ratio & Class") +
  scale_fill_brewer(palette="Pastel2")

vc <- ggplot(data = agg, aes(x = Class, y = Volume, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Volume & Class") +
  scale_fill_brewer(palette="Pastel2")

sc <- ggplot(data = agg, aes(x = Class, y = Shuck, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Shuck Weight & Class") +
  scale_fill_brewer(palette="Pastel2")

grid.arrange(cr, vc, sc, nrow=3)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
infant_vol <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="pink") + 
  ggtitle("Infant Volume & Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1100)) +
  ylab("Volume") +
  xlab("Rings")

adult_vol <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="orange") + 
  ggtitle("Adult Volume & Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1100)) +
  ylab("Volume") +
  xlab("Rings")
  
infant_weigh <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="pink") + 
  ggtitle("Infant Whole Weight & Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

adult_weigh <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="orange") + 
  ggtitle("Adult Whole Weight & Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole Weight") +
  xlab("Rings")

grid.arrange(infant_vol, adult_vol, infant_weigh, adult_weigh, nrow=2)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: This displays depict abalone growth to be most pronounced during the beginning of their lifespan. Volume and weight increase exponentially during this time. It does eventually stabilize around the 10-11 ring mark. Interestingly, there is a similarity between volume and weight distributions of adults and infants. 

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: This data reveals the difficulty in declaring the correct variables of measurement. Sex, volume, and whole weight do provide some prediction of abalone age, but there is a need for other variables that have more predictive ability. This is clearly shown as the variability of of the variables provided are quite significant and have wide distributions.

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Questions: How this sample population was determined? How was the data gathered? Is there high variability in the data? What challenges existed in measuring the data?

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: The primary issue in observational studies is lack of control, measurement errors, confounding variables, and selection bias. Validity, and results are effected by these difficulties as it can be hard to ensure certain variables are influencing the outcome and if they are measured properly. Observational studies are not able to determine causality. They merely highlight differences between groups and point towards correlated variables. However, they do offer insights to real-world trends and patterns.
