---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Russell, Jamia"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME


str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}

hist(mydata$RATIO, main="Histogram of RATIO", xlab="RATIO", col="pink")
qqnorm(mydata$RATIO, main="Q-Q Plot of RATIO")
qqline(mydata$RATIO, col="pink")
skewness(mydata$RATIO)
kurtosis(mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

L_RATIO <- log10(mydata$RATIO)
hist(L_RATIO, main="Histogram of Log 10 of RATIO", xlab="Log 10 of RATIO", col="pink")
qqnorm(L_RATIO, main="Q-Q Plot of Log 10 of RATIO")
qqline(L_RATIO, col="green")
skewness(L_RATIO)
kurtosis(L_RATIO)
mydata <- cbind(mydata,L_RATIO)
boxplot(L_RATIO ~ CLASS, mydata, notch=TRUE, main="Log 10 of RATIO by CLASS", col="pink")


```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}

bartlett.test(RATIO ~ CLASS, mydata)
bartlett.test(L_RATIO ~ CLASS, mydata)
qchisq(.95,4)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes. This is due to skewness being closer to zero. The visuals show a an even distribution and K-squared value much smaller than RATIO. RATIO is showing itself to skew right.***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

AOV1 <- aov(L_RATIO~CLASS*SEX, mydata)
summary(AOV1)
AOV2 <- aov(L_RATIO~CLASS+SEX, mydata)
summary(AOV2)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: (The non-significant interaction term did not impact L_Ratio much suggesting that the two variables CLASS and SEX are not interrelated and effect L-RATIO independently.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}

TukeyHSD(AOV2, conf.level=0.95)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: The coefficients across age classea is negative. At this confidence level, there is statistical significance across all classes. The high p-value of Male and Female allows us to assume gender is insignificant.***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
par(mfrow = c(1,2))
hist(mydata$VOLUME[mydata$TYPE=="I"], main="Volume for Infants", xlab="Infant Volume", xlim=c(0,1000), ylim=c(0,140), col="pink")
hist(mydata$VOLUME[mydata$TYPE=="ADULT"], main="Volume for Adults", xlab="Adult Volume", xlim=c(0,1000), ylim=c(0,140), col="green")
par(mfrow = c(1,1))

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The distributions show adult abalones to more normally distributed containing a wider range in volume. Infant abalones are skewed negatively to the right with a smaller range of volume. Average infant weight 100, compared to adukt 400 would make it easy for separation.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}

L_SHUCK <- log10(mydata$SHUCK)
L_VOLUME <- log10(mydata$VOLUME)
mydata <- cbind(mydata,L_SHUCK,L_VOLUME)
plot(mydata$VOLUME, mydata$SHUCK, main="SHUCK by VOLUME", col=mydata$CLASS, xlab="VOLUME", ylab="SHUCK", cex=.8)
legend(0,160,unique(mydata$CLASS),col=1:length(mydata$CLASS), pch=1, bty="n", title="CLASS", cex=.8)
plot(mydata$L_VOLUME, mydata$L_SHUCK, main="Log 10 SHUCK by Log 10 VOLUME", col=mydata$CLASS, xlab="L_VOLUME", ylab="L_SHUCK", cex=.8)
legend(0.5,2.25,unique(mydata$CLASS),col=1:length(mydata$CLASS), pch=1, bty="n", title="CLASS", cex=.8)
plot(mydata$VOLUME, mydata$SHUCK, main="SHUCK by VOLUME", col=mydata$TYPE, xlab="VOLUME", ylab="SHUCK", cex=.8)
legend(0,160,unique(mydata$TYPE),col=1:length(mydata$TYPE), pch=1, bty="n", title="TYPE", cex=.8)
plot(mydata$L_VOLUME, mydata$L_SHUCK, main="Log 10 SHUCK by Log 10 VOLUME", col=mydata$TYPE, xlab="L_VOLUME", ylab="L_SHUCK", cex=.8)
legend(0.5,2.25,unique(mydata$TYPE),col=1:length(mydata$TYPE), pch=1, bty="n", title="TYPE", cex=.8)



```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Log transformation have reduced variability. The general linear relationships have not changed,  but there is a closer fit to the regression. Shuck increase and weight increase have a strong logical trend.***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

index <- (mydata$CLASS == "A5")|(mydata$CLASS == "A4")
mydata$TYPE[index] <- combineLevels(mydata$TYPE[index], levs = c("I", "ADULT"), "ADULT")
model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data=mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: Classes A2-A5 have negative coefficients. As they mature  shuck weight will increase slowly while volume increases at a faster rate. This is reflected in the previous plots as well.WE also know that L_VOLUME has the largest predicting power due to its increased weight on the outcomes as time continues. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE is not quite significant, based on its coefficient of -0.021, compared to other variables it has less of bearing on predictions.The likelihood of predicting L_SHUCK for the harvesting decision is slim, although it is larger than that of classes.***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}

hist(model$residuals, main="Histogram of Residuals", col="pink", xlab="Residuals")
qqnorm(model$residuals, main="QQ Plot of Residuals", col="green")
qqline(model$residuals)
skewness(model$residuals)
kurtosis(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}

par(mfrow = c(2,2))
plot(mydata$L_VOLUME, model$residuals, main="Residuals by Log 10 VOLUME", col=mydata$CLASS, xlab="L_VOLUME", ylab="Residuals", cex=.8, xlim=c(0.5,4))
legend(3.5,0.35,unique(mydata$CLASS),col=1:length(mydata$CLASS), pch=1, bty="n", title="CLASS", cex=.8)
plot(mydata$L_VOLUME, model$residuals, main="Residuals by Log 10 VOLUME", col=mydata$TYPE, xlab="L_VOLUME", ylab="Residuals", cex=.8, xlim=c(0.5,4))
legend(3.1,0.3,unique(mydata$TYPE),col=1:length(mydata$TYPE), pch=1, bty="n", title="TYPE", cex=.8)
boxplot(model$residuals ~ mydata$CLASS, main="Residuals by CLASS", ylab="Residuals")
boxplot(model$residuals ~ mydata$TYPE, main="Residuals by TYPE", ylab="Residuals")
par(mfrow = c(1,1))
bartlett.test(model$residuals ~ mydata$CLASS)
qchisq(.95,4)


```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: They distributions average at 0 and are normally distributed with some give and take. Volume can be useful in harvesting decisions.***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
q6<-sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /total.adults  # [1] 0.2476573

q6

# Add code to calculate the proportion of infants harvested

q6b <- sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /total.infants

q6b

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:

med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"])

# Add code to calculate the proportion of infants harvested

q6c <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /total.infants

q6c

# Add code to calculate the proportion of adults harvested

q6c2 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /total.adults
  
q6c2

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:

med_ad_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"])

# Add code to calculate the proportion of infants harvested

q6c3 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_ad_vol) /total.infants

q6c3

# Add code to calculate the proportion of adults harvested

q6c4 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_ad_vol) /total.adults

q6c4

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}

ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="pink")) +
  geom_line(aes(volume.value, prop.adults, color="green")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+10, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") +
  ylab("Proportion") +
  scale_color_manual(values=c("pink", "green"), name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))


```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The adult median (384.56 cm^3) compared to the infant median (133.82 cm^3) is significantly larger.This suggests that the use of the adult median will preserve more abalone. This will decrease likelihood of harvesting an infant.
**

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

Note on ROC:

There are multiple packages that have been developed to create ROC curves. However, these packages - and the functions they define - expect to see predicted and observed classification vectors. Then, from those predictions, those functions calculate the true positive rates (TPR) and false positive rates (FPR) and other classification performance metrics. Worthwhile and you will certainly encounter them if you work in R on classification problems.
However, in this case, we already have vectors with the TPRs and FPRs. Our adult harvest proportion vector, (1 - prop.adults), is our TPR. This is the proportion, at each possible 'rule,' at each hypothetical harvest threshold (i.e. element of volume.value), of individuals we will correctly identify as adults and harvest. Our FPR is the infant harvest proportion vector, (1 - prop.infants). We can think of TPR as the Confidence level (ie 1 - Probability of Type I error and FPR as the Probability of Type II error. At each possible harvest threshold, what is the proportion of infants we will mistakenly harvest?
Our ROC curve, then, is created by plotting (1 - prop.adults) as a function of (1 - prop.infants). In short, how much more 'right' we can be (moving upward on the y-axis), if we're willing to be increasingly wrong; i.e. harvest some proportion of infants (moving right on the x-axis)?


-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

diff <- (1 - prop.adults) - (1 - prop.infants)
ggplot() + 
  geom_line(aes(volume.value, diff), color="pink") +
  ggtitle("Difference in Proportions Harvested") +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+20, y=.2, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+20, y=.2, angle=90) +
  xlab("Volume") +
  ylab("Difference in Proportions Harvested")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
mvol <- volume.value[which.max(smooth.difference)]
mvol

difference <- ggplot() + 
  geom_line(aes(volume.value, diff), color='orange') +
  geom_line(aes(volume.value, smooth.difference), color='yellow', linetype=2) +
  geom_vline(xintercept = mvol) +
  annotate("text", label=paste("Volume = ",round(mvol, 2)), x=mvol+40, y=.2, angle=90) +
  ggtitle("Difference in Proportions Harvested") +
  xlab("Volume") +
  ylab("Difference in Proportions Harvested")

prop <- ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="black")) +
  geom_line(aes(volume.value, prop.adults, color="blue")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  geom_vline(xintercept = mvol, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Maximum Difference = ",round(mvol, 2)), x=mvol+30, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") +
  ylab("Proportion") +
  scale_color_manual(values=c('pink','green'), name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))

grid.arrange(difference, prop, nrow=1)


```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
q7<-sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332
q7

q7d <- (1 - prop.infants)[which.max(smooth.difference)]

q7d

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

q8<-volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

q8

# Proportion of Infants
q8a <- sum(mydata[mydata$TYPE == "I", "VOLUME"] > q8) / total.infants

q8a

# Proportion of Adults
q8aa <- sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] > q8) / total.adults

q8aa
```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}

ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="pink")) +
  geom_line(aes(volume.value, prop.adults, color="green")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  geom_vline(xintercept = mvol, linetype=2) +
  geom_vline(xintercept = q8, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+10, y=.25, angle=90) + 
  annotate("text", label=paste("Maximum Difference = ",round(mvol, 2)), x=mvol+10, y=.35, angle=90) +
  annotate("text", label=paste("Zero A1 Inf = ",round(q8, 2)), x=q8+10, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") +
  ylab("Proportion") +
  scale_color_manual(values=c('pink','green'), name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}

ggplot() + 
  geom_line(aes(x = 1 - prop.infants, y = 1 - prop.adults)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  ggtitle("ROC Curve - Adults & Infants Harvested") +
  xlab("Infant Harvest Population") +
  ylab("Adult Harvest Population") +
  geom_point(aes(x = q6b, y = q6), size = 3, shape = 1) +
  geom_point(aes(x = q6c, y = q6c2), size = 3, shape = 1) +
  geom_point(aes(x = q6c3, y = q6c4), size = 3, shape = 1) +
  geom_point(aes(x = q7d, y = q7), size = 3, shape = 1) +
  geom_point(aes(x = q8a, y = q8aa), size = 3, shape = 1) +
  annotate("text", label = paste("Largest Infant = ", round(max_inf_vol, 2)), x = q6b + .17, y = q6) +
  annotate("text", label = paste("Median Infant = ", round(med_inf_vol, 2)), x = q6c + .2, y = q6c2) +
  annotate("text", label = paste("Median Adult = ", round(med_ad_vol, 2)), x = q6c3 + .17, y = q6c4) +
  annotate("text", label = paste("Maximum Difference = ", round(mvol, 2)), x = q7d + .25, y = q7) +
  annotate("text", label = paste("Zero A1 Inf = ", round(q8, 2)), x = q8a + .2, y = q8aa)


```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}

auc(1-prop.infants, 1-prop.adults)

```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	
q6_t <- (q6*total.adults + q6b*total.infants)/(total.adults+total.infants)
q6c_t <- (q6c2*total.adults + q6c*total.infants)/(total.adults+total.infants)
q6c3_t <- (q6c4*total.adults + q6c3*total.infants)/(total.adults+total.infants)
q7_t <- (q7*total.adults + q7d*total.infants)/(total.adults+total.infants)
q8_t <- (q8aa*total.adults + q8a*total.infants)/(total.adults+total.infants)

q6_1 <- c(max_inf_vol, q6, q6b, q6_t)
q6c_2 <- c(med_inf_vol, q6c2, q6c, q6c_t)
q6c_3 <- c(med_ad_vol, q6c4, q6c3, q6c3_t)
q7_4 <- c(max, q7, q7d, q7_t)
q8_5 <- c(q8, q8aa, q8a, q8_t)
c <- rbind(q6_1, q6c_2, q6c_3, q7_4, q8_5)

colnames(c) <- c("Volume", "True Posb", "False Pos", "Total Harvest")
rownames(c) <- c("Protect All Infants", "Median Infants", "Median Adults", "Max Difference", "Zero A1 Infants")
kable(c)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: The cutoffs of maximum difference between adult and infant abalone proportions shows a low true positive rate and low proportion of both types harvested. The Zero A1 Infant cutoff is the highest false positive. The cutoff representing median infant produced the highest proportions of abalones harvested, but lowest volume. PAI cutoff has the highest volume and lowest proportion harvested.***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer:  I would outline various choices and tradeoffs. Making a specific recommendation would require the input of business strategists, biologists, and other experts.Accessing the information within the assignment I would recommend the use of max difference due to the relationship between volume and proportion harvested. (2) Limitations are there is no control in the sample population which inevitably leads to issues existing throughout the analysis process that could heavily skew our findings. Outliers showed up throughout all of our plots and graphs, it would be important to include them to gain a better understanding of anomalies. (3) I would suggest harvesting earlier to prevent harvesting of infant abalone. (4) For future studies, I would suggest ensuring a random samples are collected that is reflective of different abalone environments. The data collection would reflect a broader population and allow for results to be applied with more confidence. In addition, I would implement a standardization of ring counting including technique and frequency. **